# Notation conventions {#notation}

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

\newcommand{\s}{\boldsymbol{s}}

<!-- \clearpage -->

This appendix uses the following notation conventions, which generally follows the guidance in @edwards2019:

* Greek symbols for parameters,

* the Latin/Roman alphabet for data (except $\boldsymbol{Q}$ and $\boldsymbol{H}$, which are used by convention),

* bold symbols for vectors or matrices (e.g., $\boldsymbol{\omega}$ is a vector and $\omega_{\s}$ is the value of $\boldsymbol{\omega}$ at point in space $\s$),

* $\phi$ for all distribution dispersion parameters for consistency with the code,

* $\mathbb{E}[y]$ to define the expected value (mean) of variable $y$,

* $\mathrm{Var}[y]$ to define the expected variance of the variable $y$,

* a $^*$ superscript represents interpolated or projected values as opposed to values at knot locations (e.g., $\boldsymbol{\omega}$ vs. $\boldsymbol{\omega}^*$), and

* where possible, notation has been chosen to match VAST [@thorson2019] to maintain consistency (e.g., $\boldsymbol{\omega}$ for spatial fields and $\boldsymbol{\epsilon}_t$ for spatiotemporal fields).

We include tables of all major indices (Table 1) and symbols (Table 2).

```{r indices}
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\s$", "Index for space; a vector of x and y coordinates",
  "$t$", "Index for time",
  "$g$", "Group"
)
knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, booktabs = TRUE)
```

```{r symbols, results='asis', eval=TRUE}
symbols <- tibble::tribble(
  ~Symbol, ~Code, ~Description,
  "$y$", "`y_i`", "Observed response data",
  "$\\mu$", "`mu_i`", "Mean",
  "$\\phi$", "`phi`", "A dispersion parameter for a distribution",
  "$f$", "`fit$family$link`", "Link function",
  "$f^{-1}$", "`fit$family$linkinv`", "Inverse link function",
  "$\\boldsymbol{\\beta}$", "`b_j`", "Parameter vector",
  "$\\boldsymbol{X}$", "`X_ij`", "A predictor model matrix",
  "$O_{\\s, t}$", "`offset`", "An offset variable at point $\\s$ and time $t$",

  "$\\omega_{\\s}$", "`omega_s`", "Spatial random field at point $\\s$ (knot)",
  "$\\omega_{\\s}^*$", "`omega_s_A`", "Spatial random field at point $\\s$ (interpolated)",
  "$\\zeta_{\\s}$", "`zeta_s`", "Spatially varying coefficient random field at point $\\s$ (knot)",
  "$\\zeta_{\\s}^*$", "`zeta_s_A`", "Spatially varying coefficient random field at point $\\s$ (interpolated)",
  "$\\epsilon_{\\s, t}$", "`epsilon_st`", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
  "$\\epsilon_{\\s, t}^*$", "`epsilon_st_A`", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
  "$\\delta_{\\s,t}$", "`b_t`", "AR(1) or random walk spatiotemporal deviations (knot)",
  "$\\alpha_g$", "`RE`", "IID random intercept deviation for group $g$",

  "$\\boldsymbol{\\Sigma}_\\omega$", "`-`", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "`-`", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "`-`", "Spatiotemporal random field covariance matrix",
  "$\\boldsymbol{Q}_\\omega$", "`Q_s`", "Spatial random field precision matrix",
  "$\\boldsymbol{Q}_\\zeta$", "`Q_s`", "Spatially varying coefficient random field precision matrix",
  "$\\boldsymbol{Q}_\\epsilon$", "`Q_st`", "Spatiotemporal random field precision matrix",

  "$\\sigma_\\alpha^2$", "`sigma_G`", "IID random intercept variance",
  "$\\sigma_\\epsilon^2$", "`sigma_E`", "Spatiotemporal random field marginal variance",
  "$\\sigma_\\omega^2$", "`sigma_O`", "Spatial random field marginal variance",
  "$\\sigma_\\zeta^2$", "`sigma_Z`", "Spatially varying coefficient random field marginal variance",
  "$\\kappa_\\omega$", "`kappa(0)`", "Spatial decorrelation rate",
  "$\\kappa_\\epsilon$", "`kappa(1)`", "Spatiotemporal decorrelation rate",

  "$\\rho$", "`ar1_rho`", "Correlation between random fields in subsequent time steps",
  "$\\boldsymbol{A}$", "`A`", "Sparse projection matrix to interpolate between knot and data locations",
  "$\\boldsymbol{H}$", "`H`", "2-parameter rotation matrix used to define anisotropy"
)
knitr::kable(symbols, caption = "Symbol notation, code representation (in model output or in model template code), and descriptions.", escape = TRUE)
  #  linesep = c(
  #   rep('', 7), # y, mean etc.
  #   '\\addlinespace',
  #   rep('', 7), # fields
  #   '\\addlinespace',
  #   rep('', 5), # covariance stuff
  #   '\\addlinespace',
  #   rep('', 5), # more covariance stuff
  #   '\\addlinespace',
  #   rep('', 99) # end
  # ))
```
